---
title: Working with Influence Functions 
author: Klaus Kähler Holst
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
    toc_depth: 2
    mathjax: "https://cdnjs.cloudflare.com/ajax/libs/mathjax/2.7.5/MathJax.js?config=TeX-AMS_CHTML.js"
vignette: >
  %\VignetteIndexEntry{Working with Influence Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ref.bib
---

```{r  include=FALSE }
library("lava")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

\newcommand{\pr}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\var}{\mathbb{V}\!\text{ar}}
\newcommand{\cov}{\mathbb{C}\!\ov{text}}
\newcommand{\IC}{\operatorname{IC}}

# Influence functions


Estimators that have parametric convergence rates can often be fully
characterized by their *influence function* (IF), also referred to as an influence
curve or canonical gradient. The IF allows for the direct estimation of properties of the estimator, including its asymptotic variance. Moreover,
estimates of the IF enable the simple combination and transformation of
estimators into new ones. 
This vignette describes how to estimate and manipulate IFs using the R-package `lava`.


Formally, let \(Z_1,\ldots,Z_n\) be iid, \(Z_i=(Y_{i},A_{i},W_{i})\sim P_0\), and \(\widehat{\theta}\) a
consistent estimator for the parameter \(\theta\in\mathbb{R}^p\). When
\(\widehat{\theta}\) is a *regular and asymptotic linear* (RAL) estimator, it has
a unique iid decomposition \begin{align*} \sqrt{n}(\widehat{\theta}-\theta) =
\frac{1}{\sqrt{n}}\sum_{i=1}^n \IC(Z_i; \theta) + o_p(1), \end{align*} where the
function \(\IC\) is the unique *Influence Function* s.t. \(\mathbb{E}\{\IC(Z_{i};
\theta)\}=0\) and \(\var\{\IC(Z_{i}; \theta)^{2}\}<\infty\) [@tsiatis2006semiparametric;
@vaart_1998_asymp]. The influence
function thus fully characterizes the estimator, and by the central limit
theorem it follows that the estimator converges weakly to a Gaussian distribution
\[
\sqrt{n}(\widehat{\theta}-\theta) \rightsquigarrow
\mathcal{N}(0, \var\{\IC(Z; \theta)\}),
\] where the empirical variance of the plugin estimator, \(\pr_{n}\IC(Z;
\widehat{\theta})^{\otimes 2} = \frac{1}{n}\sum_{i=1}^n \IC(Z_{i};
\widehat{\theta})\IC(Z_{i};
\widehat{\theta})^{\top}\) can be used to obtain a consistent estimate of the
asymptotic variance.

The IFs are easily derived for the parameters of many parametric statistical
models as illustrated in the [next example sections](#example-generalized-linear-model). More generally, the IF can
also be derived for a smooth target parameter \(\Psi:
\mathcal{P}\to\mathbb{R}\) where \(\mathcal{P}\) is a family of probability
distributions forming the statistical model, which often can be left completely
non-parametric. 
Formally, the parameter must be *pathwise differentiability* see [@vaart_1998_asymp]
in the sense that there exists linear bounded function  \(\dot\Psi\colon
L_{2}(P_{0})\to\mathbb{R}\) such that 
\(
[\Psi(P_{t}) - \Psi(P_{0}))]t^{-1} \to \dot\Psi(P_{0})(g) 
\)
as \(t\to 0\) for any parametric submodel \(P_t\) 
with score model \(g(z)= \partial/(\partial t) \log (p_t)(z)|_{t=0}\).  Riesz's
representation theorem then tells us that the directional derivative has a 
unique representer, \(\phi_{P_{0}}\) lying in the closure of the submodel score
space (the *tangent space*), s.t.
\begin{align*}
    \dot\Psi(P_0)(g) = \langle\phi_{P_0}, g\rangle =
    \int \phi_{P_0}(Z)g(X)\,dP_0
\end{align*}
The unique representer is exactly the IF which can be found by solving the above
integral equation.
For more details on how to derive influence
functions, we refer to [@targetedlearning_2011; @hines2022].

As an example we might be interested in the target parameter \(\Psi(P) =
\E_P(Z)\) which can be shown to have the unique (and thereby efficient)
influence function \(Z\mapsto Z-\E_P(Z)\) under the non-parametric model. 
Another target parameter could be \(\Psi_{a}(P) = \E_{P}[\E_{P}(Y\mid A=a, W)]\) which is often
a key interest in causal inference and which has the IF 
\begin{align*}
\IC(Y,A,W; P) = \frac{1(A=a)}{\pr(A=a\mid W)}(Y-\E_{P}[Y\mid A=a,W]) + \E_{P}[Y\mid
A=a,W] - \Psi_{a}(P)
    \end{align*}
See section on [average treatment effects](#average-treatment-effects).


# Examples
    
To illustrate the methods we consider data arising from the model \(Y_{ij} \sim
Bernoulli\{\operatorname{expit}(X_{ij} + A_{i} + W_{i})\}, A_{i} \sim
Bernoulli\{\operatorname{expit}(W_{i})\}\) 
with independent covariates
\(X_{ij}\sim\mathcal{N}(0,1), U_{i}\sim\mathcal{N}(0,1)\). 
```{r sim_model}
m <- lvm() |>
    regression(y1 ~ x1 + a + w) |>
    regression(y2 ~ x2 + a + w) |>
  regression(y3 ~ x3 + a + w) |>
  regression(a ~ w) |>
  distribution(~ y1 + y2 + y3 + a, value = binomial.lvm()) |>
  distribution(~id, value = Sequence.lvm(integer = TRUE))
```

We simulate from the model where \(Y_3\) is only observed for half of the subjects
```{r simulate}
n <- 5e2
dw <- sim(m, n, seed=1) |>
  transform(z = y3 * ifelse(id > n / 2, NA, 1))
dl <- mets::fast.reshape(dw, varying="y")
lava:::Print(dl)
```

## Example: population mean

The main functions for working with influence functions are 

- `estimate` which prepares a model object and estimates the IF and
  corresponding robust standard errors. Can also be used to transform model
  parameters by application of the Delta Theorem.
- `merge` method for combining estimates via their estimated IFs
- `IC` method to extract the estimated IF

Here we first consider the problem of estimating the IF of the mean. For the
simulated data let
\(\widehat{p}_1\) denote the MLE of the proportion
\(\pr(Y_{1}=1)\), it then follows that
\[
\sqrt{n}\{\widehat{p}_1 - \pr(Y_{1}=1)\} = \frac{1}{\sqrt{n}}\sum_{i=1}^{n}
1(Y_{1i}=1) - \pr(Y_{1}=1)
\] 
and hence the IF is given by \(1(Y_{1i}=1) - \pr(Y_{1}=1)\). 

To estimate this parameter and its IF we will use the `estimate` function

```{r}
inp <- as.matrix(dw[, c("y1", "y2", "y3")])
e <- estimate(inp[, 1, drop = FALSE])
e
```
The reported standard errors from the `estimate` method
are the robust standard errors obtained from the IF.
    The variance estimate and the parameters can be extracted with the `vcov`
    and `coef`methods.
The IF itself can be extracted with the `IC` method:
```{r}
IC(e) |> Print()
```

It's also possible to simultaneously estimate the proportions of each of the three
binary outcomes 
```{r}
estimate(inp)
```
or alternatively the input can be a model object, here a `mlm` object:
```{r}
e <- lm(cbind(y1, y2, y3) ~ 1, data = dw) |>
  estimate()
IC(e) |> head()
```

The IF for both the empirical mean and variance can also be estimated directly with the `IC` method:
```{r}
with(dw, IC(y1)) |> head()
```
with the point estimates being stored in the attributes "mean" and "variance".

## Example: generalized linear model

For a \(Z\)-estimator defined by the score equation \(E[U(\theta; Z)] = 0\), the
IF is given by \begin{align*} IC(Z; \theta) =
\E\Big\{\frac{\partial}{\partial\theta^\top}U(\theta; Z)\Big\}^{-1}U(\theta; Z)
\end{align*}
In particular, for a MLE the score, \(U\), is given by the partial derivative of
the log-likelihood function.

As an example, we can obtain the estimates with robust standard errors for a
logistic regression model:
```{r glm}
g1 <- glm(y1 ~ a + x1, data = dw, family = binomial)
estimate(g1)
```
The IF can be extracted from the `estimate` object or directly from the 
model object
```{r}
IC(g1) |> head()
```


The same estimates can be obtained with a 
*cumulative link regression* model which also generalizes to 
ordinal outcomes
```{r ordreg}
ordreg(y1 ~ a + x1, dw, family=binomial(logit)) |> estimate()
```

## Example: right-censored outcomes

To illustrate the methods on survival data we will use the Mayo Clinic Primary
Biliary Cholangitis Data [@therneau00surv]

```{r mets}
library("survival")
data(pbc, package="survival")
```

The Cox proportional hazards model can be fitted with the `mets::phreg` method
which can estimate the IF for both the partial likelihood parameters and the
baseline hazard. Here we fit a survival model with right-censored event times
```{r phreg}
fit.phreg <- mets::phreg(Surv(time, status > 0) ~ age + sex, data = pbc)
fit.phreg
IC(fit.phreg) |> head()
```

The IF for the baseline cumulative hazard at a specific time point can be
estimated in similar way:
```{r phreg-baseline}
baseline <- function(object, time, ...) {
  ic <- IC(object, baseline = TRUE, time = time, ...)
  est <- mets::predictCumhaz(object$cumhaz, new.time = time)[1, 2]
  estimate(NULL, coef = est, IC = ic, labels = paste0("chaz:", time))
}
tt <- 2000
baseline(fit.phreg, tt)
```

The `estimate` and `IF` methods are also available for parametric survival
models via `survival::survreg`, here a Weibull model:
```{r survreg}
survival::survreg(Surv(time, status > 0) ~ age + sex, data = pbc, dist="weibull") |>
  estimate()
```

<!-- glm poisson -->
<!-- ```{r lifetable} -->
<!-- pbc2 <- mets::lifetable(Surv(time, status > 0) ~ sex, data = pbc, breaks = seq(0, 5000, length.out = 10)) -->
<!-- head(pbc2) -->
<!-- glm(events ~ offset(log(atrisk)) + sex + factor(int.end), data = pbc2, family=poisson(log)) -->
<!-- ``` -->

# Combining influence functions

merge, +, id, +.estimate, 

# Transformations and the Delta theorem

estimate(,...)

# Clustered data & GEE

id

##' ## Clusters and subset (conditional marginal effects)
##' d$id <- rep(seq(nrow(d)/4),each=4)
##' estimate(g,function(p,data)
##'          list(p0=lava:::expit(p[1] + p["z"]*data[,"z"])),
##'          subset=d$z>0, id=d$id, average=TRUE)
##'
##' ## More examples with clusters:
##' m <- lvm(c(y1,y2,y3)~u+x)
##' d <- sim(m,10)
##' l1 <- glm(y1~x,data=d)
##' l2 <- glm(y2~x,data=d)
##' l3 <- glm(y3~x,data=d)
##'
##' ## Some random id-numbers
##' id1 <- c(1,1,4,1,3,1,2,3,4,5)
##' id2 <- c(1,2,3,4,5,6,7,8,1,1)
##' id3 <- seq(10)
##'
##' ## Un-stacked and stacked i.i.d. decomposition
##' IC(estimate(l1,id=id1,stack=FALSE))
##' IC(estimate(l1,id=id1))
##'
##' ## Combined i.i.d. decomposition
##' e1 <- estimate(l1,id=id1)
##' e2 <- estimate(l2,id=id2)
##' e3 <- estimate(l3,id=id3)
##' (a2 <- merge(e1,e2,e3))
##'
##' ## If all models were estimated on the same data we could use the
##' ## syntax:
##' ## Reduce(merge,estimate(list(l1,l2,l3)))
##'
##' ## Same:
##' IC(a1 <- merge(l1,l2,l3,id=list(id1,id2,id3)))
##'
##' IC(merge(l1,l2,l3,id=TRUE)) # one-to-one (same clusters)
##' IC(merge(l1,l2,l3,id=FALSE)) # independence

# Hypothesis testing

pairwise, -, estimate, compare, var_ic

multiple testing

pzmax,p.correct, closed testing procedure

##' ## Testing contrasts
##' estimate(g,null=0)
##' estimate(g,rbind(c(1,1,0),c(1,0,2)))
##' estimate(g,rbind(c(1,1,0),c(1,0,2)),null=c(1,2))
##' estimate(g,2:3) ## same as cbind(0,1,-1)
##' estimate(g,as.list(2:3)) ## same as rbind(c(0,1,0),c(0,0,1))
##' ## Alternative syntax
##' estimate(g,"z","z"-"x",2*"z"-3*"x")
##' estimate(g,z,z-x,2*z-3*x)
##' estimate(g,"?")  ## Wildcards
##' estimate(g,"*Int*","z")
##' estimate(g,"1","2"-"3",null=c(0,1))
##' estimate(g,2,3)

##' ## Usual (non-robust) confidence intervals
##' estimate(g,robust=FALSE)
##'
##' ## Transformations
##' estimate(g,function(p) p[1]+p[2])
##'
##' ## Multiple parameters
##' e <- estimate(g,function(p) c(p[1]+p[2],p[1]*p[2]))
##' e
##' vcov(e)


```{r}
m <- categorical(lvm(),~x,K=5)
regression(m,additive=TRUE) <- y~x
d <- simulate(m,100,seed=1,'y~x'=0.1)
l <- lm(y~-1+factor(x),data=d)

f <- function(x) coef(lm(x~seq_along(x)))[2]
null <- rep(mean(coef(l)), length(coef(l)))
## just need to make sure we simulate under H0: slope=0
## estimate(l,f,R=1e2,null.sim=null)
## estimate(l,f)
```



# Averaging

Let \(Z_{1},\ldots,Z_{n}\) be iid observations, \(Z_{1}\sim P\) and let  \(X_{i}\subset Z_{i}\).

Assume that \(\widehat{\theta}\) is RAL estimator of \(\theta\in\Omega\subset \mathbb{R}^{p}\)
\[\sqrt{n}(\widehat{\theta}-\theta) = \frac{1}{\sqrt{n}}\sum_{i=1}^{n}\phi(Z_{i}) + o_{p}(1).
\]
Let \(f:\mathcal{X}\times\Omega\to\mathbb{R}\) be continuous differentiable in \(\theta\)
\[\sqrt{n}\{f(X; \widehat{\theta})-f(X; \theta)\} =
\frac{1}{\sqrt{n}}\nabla_{\theta}f(X;\theta)\sum_{i=1}^{n}\phi(Z_{i}) + o_{p}(1).
\]


Let \(\Psi = Pf(X;\theta)\) and \(\widehat{\Psi} = P_{n} f(X;\widehat{\theta})\). \(P\) and \(P_{n}\) are here everywhere the integrals wrt. \(X\). It is easily verified that
\begin{align*}
\widehat{\Psi}-\Psi &= (P_{n}-P)(f(X; \theta)-\Psi) + P[f(X;\widehat{\theta})-f(X;\theta)] \\
&\quad + (P_{n}-P)[f(X;\widehat{\theta})-f(X;\theta)]
\end{align*}

From Lemma 19.24 [@vaart_1998_asymp] it follows that for the last term
\begin{align*}
\sqrt{n}(P_{n}-P)[f(X;\widehat{\theta})-f(X;\theta)] = o_{p}(1)
\end{align*}
when \(f\) for example is Lipschitz and more generally when \(f(X;\theta)\) forms a \(P\)-Donsker class.

It therefore follows that
\begin{align*}
\sqrt{n}(\widehat{\Psi}-\Psi) &= \sqrt{n}P_{n}(f(X; \theta)-\Psi) +
\frac{1}{\sqrt{n}}P\nabla_{\theta}f(X;\theta)\sum_{i=1}^{n}\phi(Z_{i}) + o_{p}(1) \\
&=
\frac{1}{\sqrt{n}}\sum_{i=1}^{n}\{f(X;\theta)-\Psi\} +
\frac{1}{\sqrt{n}}P\nabla_{\theta}f(X;\theta)\sum_{i=1}^{n}\phi(Z_{i}) + o_{p}(1)
\end{align*}
Hence the IF becomes
\[
IC(Z) = f(X;\theta)-\Psi +
[P\nabla_{\theta}f(X;\theta)]\phi(Z).
\]

estimate(average=TRUE)

##' ## Marginalize
##' f <- function(p,data)
##'   list(p0=lava:::expit(p["(Intercept)"] + p["z"]*data[,"z"]),
##'        p1=lava:::expit(p["(Intercept)"] + p["x"] + p["z"]*data[,"z"]))
##' e <- estimate(g, f, average=TRUE)
##' e
##' estimate(e,diff)
##' estimate(e,cbind(1,1))


## Average Treatment Effects

```{r}
g <- glm
```


targeted::cate
targeted::riskreg
targeted::alean

# Stacking 
stack

## Measurement error
measurement.error

# Examples diagnostics and ....

diagtest, assoc, riskcomp, Ratio, Diff, odds, OR,
information_assoc, information.table, independence (not exported) information.multinomial
gkgamma, kappa


# estimate methods

coef, vcov, summary, parameter

level

coef.summary.estimate

estimate.list
estimate.array
multinomial



# Enabling new models 

IC method
score method

# SessionInfo

```{r sessionInfo}
sessionInfo()
```

# Bibliography

